Bayesian spectral density estimation using P-splines with quantile-based knot placement

### 2020-10-31 response.tex

\documentclass{scrartcl}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{color}
\usepackage[english]{babel}
%\usepackage[fixlanguage]{babelbib}
\usepackage{natbib}
\usepackage[pdftex]{graphicx}
\usepackage{dsfont}
\usepackage{color}
\usepackage{hyperref}  % Package for hyperlinking
%\usepackage{a4wide}
\usepackage[a4paper, left=3cm, right=3cm, top=2cm]{geometry}
%\usepackage{texititel}
\title{Bayesian spectral density estimation using P-splines with quantile-based knot placement}
\subtitle{Response letter to Statistical Papers}
\author{Patricio Maturana-Russel, Renate Meyer}

@@

\pagestyle{empty}
\begin{document}
\maketitle\thispagestyle{empty}

% -----------------------------------------------------------------------------
% -----------------------------------------------------------------------------
% -----------------------------------------------------------------------------

\subsection*{Referee \#1}

\response{We would like to thank the referee for a thorough review of our paper. Below, you find a point-by-point response to your points raised. We have now carefully revised the paper taking all of your -- and the other referee's -- comments into account and hope that you will now find the paper acceptable for publication. For your convenience, all changes made are marked in red in the manuscript.}

\textbf{Major comment:}

\refereeQuote{p.8, Quantile-based knot placement\\
I could not understand the meaning of ``calculate its distribution function $F_z = P(z \leq z_i)$" in the second item of the procedure.  The values of the transformed periodogram $z_i$ are regarded as the values of the density $\tilde{f}$, which is non-monotonic, taken at equally-spaced frequencies.  According to item 4, the distribution function $F_z$, and hence the density function $\tilde{f}$, should be defined on the frequency domain. \\
I had a look at the R codes of the package psplinePsd. I think that the density function $\tilde{f}$ is a linear spline that interpolates as
\begin{align*}
\tilde{f} \left(\dfrac{\lambda_k}{2\pi}\right) = \dfrac{z_k}{\sum_{i=1}^{\nu}z_i}, \quad k=1,\dots,\nu.
\end{align*}
}

\response{That's correct!  The $\widetilde{f}$ is actually a probability mass function (not a probability density function), which is interpolated.  However, we have simplified a bit our knot allocation algorithm avoiding that step.  The new version of the algorithm is described step by step in our manuscript.  This change leads to small changes in our application section, which has been consequently updated.  We have also updated the \texttt{psplinePsd} package.}\bigskip

\textbf{Minor comments:}
\begin{enumerate}
\item
\refereeQuote{p.4, line 10\\
A spline of degree $r+1\implies$ A spline of degree $r$ (or, of order $r+1$)}\smallskip

\response{Fixed.}\bigskip

\item
\refereeQuote{p.5 line 13\\
In the definition of $\tau, f(\tau \omega)\implies f(\pi \omega)$}\smallskip

\response{That is correct.  Fixed. }\bigskip

@@

\end{enumerate}

% -----------------------------------------------------------------------------
% -----------------------------------------------------------------------------
% -----------------------------------------------------------------------------

\subsection*{Referee \#2}

\response{We would like to thank the referee for a thorough review of our paper. Below, you find a point-by-point response to your points raised. We have now carefully revised the paper taking all of your -- and the other referee's -- comments into account and hope that you will now find the paper acceptable for publication. For your convenience changes are marked in red in the manuscript.
} \smallskip

\textbf{Specific comments:}
\begin{enumerate}
	\item \refereeQuote{ Page 5, how to choose penalty or prior $\phi$?}

	\response{In a Bayesian setting, the roughness penalty is translated into the prior distribution for the spline parameters via a normal distribution.  See for instance \cite{Bremhorst:2016}.  In order to make the posterior less sensitive to the prior, we suggest to follow their proposal and use their robust specifications for the hierarchical priors on $\phi$, which is also described in our paper. }

	\item \refereeQuote{To select the number of B-spline densities, the authors employ a heuristic rule of thumb $K^*$ in the application Section, it is of interest to compare it with other criteria such as AIC and BIC.}\smallskip

	\response{This could be a potential avenue for future research, however, we think it is beyond the scope of this paper.  An analysis of those characteristics can be quite extensive and could constitute a paper in itself, see for instance \cite{Likhachev2017}.  We would like to highlight that the number of knots is considered non-important by many, since the penalty parameter plus the penalty order difference penalties specified in the penalty matrix $\textbf{P}$ penalise the inclusion of B-spline densities, controlling the smoothness of the estimate.  This means that we only need to select a minimum value, the penalty terms take care of the rest. To quote \cite{Eilers2015}: ``With the number of B-splines in the basis we can tune the smoothness of a curve to
the data at hand. A smaller number of splines gives a smoother result. However, this is not the only possibility. We can also use a large basis and additionally constrain the
coefficients of the B-splines, to achieve as much smoothness as desired. A properly
chosen penalty achieves this"; and also ``The number of B-splines can be (much) larger than the number of observations. The
penalty makes the fitting procedure well-conditioned. This should be taken literally:
even a thousand splines will fit ten observations without problems. Such is the power
of the penalty."}\bigskip

	Other minor comments: \bigskip

	\item \refereeQuote{Page 2, line 17, what is $\lambda_i$?}\smallskip

	\response{$\lambda_i$ denote the Fourier frequencies. These have now been defined just after the first equation on page 2.} \bigskip

	\item \refereeQuote{ Page 4, lines 5-7, ``In particular, the approximation error of the mixture of B-spline distributions can be made arbitrarily small by finding suitable knot locations and increasing the number of knots", what does ``made arbitrarily small" mean? Do you mean the approach with the beta densities the error cannot be made arbitrarily small?}\smallskip

	\response{Arbitrarily small means that the approximation error converges to zero with increasing number of knots. We have  cited the paper by \cite{Perron:2001} that make this statement precise. But we do not mean to imply that this is not the case for the mixture of beta densities. Of course, according to the well-known Weierstrass approximation theorem, any continuous function can be approximated in the supnorm by a linear combination of Bernstein polynomials. If the continuous function is a density, the approximant becomes a mixture of beta distributions. By increasing the number of beta distributions, the approximation error can be made arbitrarily small. But as shown in Perron and Mengersen (2001) the rate of approximation is faster for Bsplines than for Bernstein polynomials. We have changed this paragraph in the manuscript accordingly.}

%\textcolor{red}{I would change the text in the paper to:
%" As shown by Perron and Mengersen (2001) for distributions on the interval [0,1], both mixtures of B-spline and beta distributions can approximate these arbitrarily well by increasing the number of mixture components, i.e.\ increasing the polynomial order for the beta distributions and the number of knots for suitable know locations for the B-splines. But the rate of approximation is faster for B-splines than for beta distributions.}
%	\response{Yes, the loss associated to the approximation given by the beta densities with respect to a loss function cannot be made arbitrarily small. This was shown by %\cite{Perron:2001} who also showed that if one replaces the beta densities by B-spline densities of fixed order but with variable knots, the loss can be made arbitrarily small by increasing %the number of knots.  We have added the corresponding reference to our manuscript.} \bigskip

	\item \refereeQuote{Page 4, line 15, ``A spline of degree r + 1 is a function", I think degree" should be order", we use degree for polynomials.}\smallskip

	\response{Fixed.} \bigskip

	\item \refereeQuote{Page 5, line 20, ``If $\tau = \int_0^1f(\tau \omega)d \omega$ is the normalizing constant", it implies $s_r$ is a density, is it correct?}

	\response{Yes, $s_r$ is a pdf.  Note that it is actually a mixture of B-spline densities.}
\end{enumerate}

\subsection*{Further Changes}
Here some extra changes:
\begin{enumerate}
	\item We have defined $K^*$ as the number of internal knots (B-splines and BspDP prior section).
	\item We have specified the prior on $\tau$ (P-spline prior subsection of section 2). 
	\item We have added $s_r(\omega)$ in Eq(3), short notation used in the posterior distribution of~$\tau$ (Posterior computation section).	
	\item We have slightly simplified our knot location approach.  This has led to minor variations in the results shown in the application section, which has been updated.  This change has not impacted the conclusions.  This version of the knot location algorithm is described step by step in the Quantile-based knot placement section.
\end{enumerate}

\bibliographystyle{plainnat}
\bibliography{pSplinesReferences}  % Name of BibTeX file

%\begin{thebibliography}{38}
%	\providecommand{\natexlab}[1]{#1}
%	\providecommand{\url}[1]{\texttt{#1}}
%	\expandafter\ifx\csname urlstyle\endcsname\relax
%	\providecommand{\doi}[1]{doi: #1}\else
%	\providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
%	\bibitem[Bremhorst and Lambert(2016)]{Bremhorst:2016}
%	Vincent Bremhorst and Philippe Lambert.
%	\newblock Flexible estimation in cure survival models using {B}ayesian
%	p-splines.
%	\newblock \emph{Computational Statistics \& Data Analysis}, 93:\penalty0 270 --
%	284, 2016.
%	\newblock ISSN 0167-9473.
%	\newblock \doi{https://doi.org/10.1016/j.csda.2014.05.009}.
%	\newblock URL
%	\url{http://www.sciencedirect.com/science/article/pii/S0167947314001492}.
%\end{thebibliography}

### 2020-10-31 pSplines.tex
Add files via upload

As the periodogram fluctuates around the true spectral density, most  nonparametric methods for spectral density estimation hinge on  smoothing the periodogram
\begin{align*}
I_n(\lambda) = \dfrac{1}{2 \pi n} \left| \sum_{t=1}^{n} Y_t \exp \left( -i t \lambda\right)\right|^2,
\end{align*}
in one way or another, where $-\pi < \lambda \leq \pi$ is the angular frequency and $\bm{Y}=(Y_1,\ldots,Y_n)$ is a stationary time series of length $n$. {\pmr The periodogram ordinates evaluated at the Fourier frequencies $\lambda \in [-\pi,\pi]$  are asymptotically independent and follow exponential distributions with mean $f(\lambda)$.  Thus, they are unbiased estimates of $f(\lambda)$  but not consistent \citep{Brockwell:1986}.}

Most Bayesian approaches to psd estimation use the so-called Whittle likelihood, introduced by \cite{Whittle:1957} as an approximation to the likelihood of a Gaussian stationary time series.  Based on the  limiting independent exponential distribution of the periodogram values, the Whittle likelihood for a mean centred weakly stationary time series $\bm{Y}$ of length $n$ is defined as
\begin{align}
L(\bm{Y} | f) &= \prod_{l=1}^{\nu}\dfrac{1}{f(\lambda_{l})} e^{-I_n(\lambda_l)/f(\lambda_l)}  \nonumber \\
&\propto \exp \left\{ -\! \sum_{l=1}^{\nu} \Bigg( \log f(\lambda_l) + \dfrac{I_{n}(\lambda_l)}{f(\lambda_l)} \Bigg) \right\},\label{eq:whittle_like}
\end{align}
where $\lambda_l = 2\pi l / n$ are the Fourier frequencies, $f(\cdot)$ spectral density of $\bm{Y}$, and $\nu = \left\lfloor (n-1)/2 \right\rfloor$ is the greatest integer value less than or equal to $(n-1)/2$. Advantages of this approximation are that it depends directly on the spectral density, unlike the true Gaussian likelihood function, and gives a good approximation for non-Gaussian time series under certain conditions as shown by \cite{ShaoXiaofeng2007ASTf}.

@@

Our knot location proposal works as follow.  We calculate the periodogram and apply a square root transformation in order to have more regular magnitudes and eliminate a potential trend.  Second, we standardize these values, apply the absolute value function,{\pmr and normalise them}.  We treat this transformed periodogram as a {\pmr mass probability} function (pmf) and interpolate its distribution function $\widetilde{F}$.  Finally, the knots are allocated according to the quantiles of the inverse distribution function $\widetilde{F}^{-1}$, {\pmr which is approximated by interpolation.}  Thus, in those areas in which the periodogram has sharp and abrupt changes, our procedure will assign more knots in proportion to their magnitudes. Hereafter, we will refer to this scheme as {\em  quantile-spaced (Q-spaced) knots}.%\textit{periodogram-spaced knots} or short \textit{P-spaced knots}.
%REMARK: HOW ABOUT "PERIODOGRAM-SPACED KNOTS"	OR 'P-SPACED KNOTS'  INSTEAD OF DISTRIBUTED KNOTS?

@@

{\pmr
The procedure for the frequencies in the unit interval $\omega_l = 2 l/n  \in[0,1]$, which can be easily translated to the frequency domain, can be summarized as follows:	
\begin{enumerate}
	\item Let $\bm{x} = \left (\sqrt{I_n(\omega_1)}, \dots, \sqrt{I_n(\omega_{\nu})} \right)$ be the square root of the periodogram values.
	\item Define $y_l = \left|\dfrac{x_l - \bar{x}}{S_x} \right|$ where $\bar{x}$ denotes the average and $S_x$ the standard deviation of ${\bm x}$.
	\item Define $z_l = \frac{y_l}{\sum_{l=1}^{\nu} y_l}$, calculate its cumulative distribution $\widetilde{F}$, and interpolate it.
	\item Let $\bm{q} = \left(q_1,\dots,q_{K^*} \right)$ be a vector of $K^*$ quantiles, equally spaced in $[0,1]$.
	\item The Q-spaced internal knot vector is given by $\bm{\xi}^* = \left(\widetilde{F}^{-1}(q_1),\dots, \widetilde{F}^{-1}(q_{K^*}) \right)$. %where the inverse function is obtained by interpolation.	
\end{enumerate}

@@

The theoretical spectral density for an AR($p$) model is given by
\begin{align*}
f(\lambda) = \dfrac{\sigma^2}{2 \pi} \dfrac{1}{|1-\sum_{j=1}^{p}\rho_j \exp(-i \lambda)|^2},
\end{align*}
where $\sigma^2$ is the variance of the white noise innovations and ($\rho_1, \dots, \rho_p$) are the model parameters.  In order to compare the psd estimates  to the true function, we use the integrated absolute error (IAE) or $L_1$-error which is defined as
\begin{align*}
\text{IAE} = || \widehat{f} - f ||_1 = \int_{0}^{\pi} |\widehat{f}(\omega)-f(\omega)|\text{d}\omega,
\end{align*}
where $\widehat{f}(\cdot)$ is the pointwise posterior median of the spectral density $f(\cdot)$.   The IAE is computed for each of the 300 replications and its median is calculated.  
The results are displayed in Table~\ref{table:sim_IAE}.

>>>>

As the periodogram fluctuates around the true spectral density, most  nonparametric methods for spectral density estimation hinge on  smoothing the periodogram
\begin{align*}
I_n(\lambda_l) = \dfrac{1}{2 \pi n} \left| \sum_{t=1}^{n} Y_t \exp \left( -i t \lambda_l\right)\right|^2,
\end{align*}
in one way or another, where $ \lambda_l=2\pi/n$ for $l=0,\ldots,\nu = \left\lfloor (n-1)/2 \right\rfloor$ are the Fourier frequencies and $\bm{Y}=(Y_1,\ldots,Y_n)$ is a stationary time series of length $n$. {\pmr The periodogram ordinates evaluated at the Fourier frequencies  are asymptotically independent and follow exponential distributions with mean $f(\lambda_l)$.  Thus, they are asymptotically  unbiased estimates of $f(\lambda_l)$  but not consistent \citep{Brockwell:1986}.}

Most Bayesian approaches to psd estimation use the so-called Whittle likelihood, introduced by \cite{Whittle:1957} as an approximation to the likelihood of a Gaussian stationary time series.  Based on the  limiting independent exponential distribution of the periodogram values, the Whittle likelihood for a mean centred weakly stationary time series $\bm{Y}$ of length $n$ is defined as
\begin{align}
L(\bm{Y} | f) &= \prod_{l=1}^{\nu}\dfrac{1}{f(\lambda_{l})} e^{-I_n(\lambda_l)/f(\lambda_l)}  \nonumber \\
&\propto \exp \left\{ -\! \sum_{l=1}^{\nu} \Bigg( \log f(\lambda_l) + \dfrac{I_{n}(\lambda_l)}{f(\lambda_l)} \Bigg) \right\}.\label{eq:whittle_like}
\end{align}
%where $\lambda_l = 2\pi l / n$ are the Fourier frequencies, $f(\cdot)$ spectral density of $\bm{Y}$, and $\nu = \left\lfloor (n-1)/2 \right\rfloor$ is the %greatest integer value less than or equal to $(n-1)/2$.
 Advantages of this approximation are that it depends directly on the spectral density, unlike the true Gaussian likelihood function, and gives a good approximation for non-Gaussian time series under certain conditions as shown by \cite{ShaoXiaofeng2007ASTf}.

@@

Our knot location proposal works as follow.  We calculate the periodogram and apply a square root transformation in order to have more regular magnitudes and eliminate a potential trend.  Second, we standardize these values, apply the absolute value function,{\pmr and normalise them}.  
We treat this transformed periodogram as a {\pmr probability mass} function (pmf) and interpolate its distribution function \textcolor{red}{ by the continuous cumulative distribution function} $\widetilde{F}$.
Finally, the knots are allocated according to quantiles of $\widetilde{F}$.
 Thus, in those areas in which the periodogram has sharp and abrupt changes, our procedure will assign more knots in proportion to their magnitudes. Hereafter, we will refer to this scheme as {\em  quantile-spaced (Q-spaced) knots}.%\textit{periodogram-spaced knots} or short \textit{P-spaced knots}.
%REMARK: HOW ABOUT "PERIODOGRAM-SPACED KNOTS"	OR 'P-SPACED KNOTS'  INSTEAD OF DISTRIBUTED KNOTS?

@@

{\pmr
The procedure for the frequencies in the unit interval $\omega_l = 2 l/n  \in[0,1]$, which can be easily translated to the frequency domain, can be summarized as follows:	
\begin{enumerate}
	\item Let $\bm{x} = \left (\sqrt{I_n(\pi\omega_1)}, \dots, \sqrt{I_n(\pi\omega_{\nu})} \right)$ be the square root of the periodogram values.
	\item Define $y_l = \left|\dfrac{x_l - \bar{x}}{S_x} \right|$ where $\bar{x}$ denotes the average and $S_x$ the standard deviation of ${\bm x}$.
	\item Define $z_l = \frac{y_l}{\sum_{l=1}^{\nu} y_l}$, $Z_l= \sum_{j\leq l} z_j$, and let $\widetilde{F}$ denote the continuous cdf that interpolates the $Z_l$, $l=1,\ldots,\nu$.
	\item Let $\bm{q} =\left( \frac{1}{K^*+1}, \frac{2}{K^*+1}\dots, \frac{K^*}{K^*+1} \right)$ be a vector of $K^*$ equally spaced points  in  $[0,1]$.
	\item The Q-spaced internal knot vector is given by the corresponding quantiles $\bm{\xi}^* = \left(\widetilde{F}^{-1}(q_1),\dots, \widetilde{F}^{-1}(q_{K^*}) \right)$. %where the inverse function is obtained by interpolation.	
\end{enumerate}

@@

The theoretical spectral density for an AR($p$) model is given by
\begin{align*}
f(\lambda) = \dfrac{\sigma^2}{2 \pi} \dfrac{1}{|1-\sum_{j=1}^{p}\rho_j \exp(-i \lambda)|^2},
\end{align*}
where $\sigma^2$ is the variance of the white noise innovations and ($\rho_1, \dots, \rho_p$) are the model parameters.  In order to compare the psd estimates  to the true function, we use the integrated absolute error (IAE) or $L_1$-error which is defined as
\begin{align*}
\text{IAE} = || \widehat{f} - f ||_1 = \int_{0}^{\pi} |\widehat{f}(\textcolor{red}{\lambda})-f(\textcolor{red}{\lambda})|\text{d}\textcolor{red}{\lambda},
\end{align*}
where $\widehat{f}(\cdot)$ is the pointwise posterior median of the spectral density $f(\cdot)$.   The IAE is computed for each of the 300 replications and its median is calculated.  
The results are displayed in Table~\ref{table:sim_IAE}.
